---
title: "AMV_chapter"
format: html
editor: visual
---

# PART 1: PCA

## Mathématiques

## Interprétation

This chapter is a simple example using R

You can import R package using the code

```{r}
library(tidyverse)
```

and then describe the purpose of your chapter as well as executing R command.

For example a basic summary of a dataset is given by

```{r}
hippo <- read.table(data="", sep = "," , header = TRUE)
```

and produce a graph

```{r}
df %>% ggplot() +
	aes(x=species, y = body_mass_g) +
	geom_boxplot()  
```

A citation @bauer2023writing

# PART 2: MCA

## Mathématiques

## Interpretation

Let's have to look to fictive the data set we will be working on:

```{r}
library(tidyverse)
hippo <- read.table(file = "https://raw.githubusercontent.com/AnsaldiL/MODE_reproduciblescience/master/hp.csv", sep=';', header=TRUE)
hippo=hippo[,-1]
str(hippo)
```

Water consumption behaviour of Hippopotamus was observed in Penjari National Park, Benin. Drinking frequency was evaluated by a technician and rated "rarely" or "regularly". Sex is indicated with F for female and M for male. As there are 3 lakes in the park, the tree of them are written "G1", "G2" and "G3".

```{r}
library(FactoMineR)
res.mca = MCA(hippo, quanti.sup=1,2)
plot.MCA(res.mca, invisible=c("var","quali.sup"), cex=0.7)
plot.MCA(res.mca, invisible=c("ind","quali.sup"), cex=0.7)
plot.MCA(res.mca, invisible=c("ind"))
plot.MCA(res.mca, invisible=c("ind", "var"))
```

## Take Home Message

MCA is a statistical method adapted to table of type "individuals x quatitative variable". Eingen Values correspond to means of squered correlation ratios. It could be used a pre-processing before a classification or a coinertia analysis on tables with quantitative.

# PART 3: RDA

## Mathématiques

## Interprétation

This chapter is a simple example using R

You can import R package using the code

```{r}
library(tidyverse)
```

and then describe the purpose of your chapter as well as executing R command.

For example a basic summary of a dataset is given by

```{r}
df <- read.table("https://gist.githubusercontent.com/slopp/ce3b90b9168f2f921784de84fa445651/raw/4ecf3041f0ed4913e7c230758733948bc561f434/penguins.csv", sep = "," , header = TRUE)
```

and produce a graph

```{r}
df %>% ggplot() +
	aes(x=species, y = body_mass_g) +
	geom_boxplot()  
```

A citation @bauer2023writing

# PART 4: RLQ

## Mathématiques

## Interprétation

This chapter is a simple example using R

You can import R package using the code

```{r}
library(tidyverse)
```

and then describe the purpose of your chapter as well as executing R command.

For example a basic summary of a dataset is given by

```{r}
df <- read.table("https://gist.githubusercontent.com/slopp/ce3b90b9168f2f921784de84fa445651/raw/4ecf3041f0ed4913e7c230758733948bc561f434/penguins.csv", sep = "," , header = TRUE)
```

and produce a graph

```{r}
df %>% ggplot() +
	aes(x=species, y = body_mass_g) +
	geom_boxplot()  
```

A citation @bauer2023writing
