---
title: "AMV_chapter"
format: html
editor: visual
---

# PART 1: PCA

## Mathématiques

## Interprétation

This chapter is a simple example using R

You can import R package using the code

```{r}
library(tidyverse)
```

and then describe the purpose of your chapter as well as executing R command.

For example a basic summary of a dataset is given by

```{r}
hippo <- read.table(data="", sep = "," , header = TRUE)
```

and produce a graph

```{r}
df %>% ggplot() +
	aes(x=species, y = body_mass_g) +
	geom_boxplot()  
```

A citation @bauer2023writing

# PART 2: MCA

## Mathématiques

## Interpretation

Let's have to look to fictive the data set we will be working on:

```{r}
library(tidyverse)
hippo <- read.table(file = "https://raw.githubusercontent.com/AnsaldiL/MODE_reproduciblescience/master/hp.csv", sep=';', header=TRUE)
hippo=hippo[,-1]
str(hippo)
```

Water consumption behaviour of Hippopotamus was observed in Penjari National Park, Benin. Drinking frequency was evaluated by a technician and rated "rarely" or "regularly". Sex is indicated with F for female and M for male. As there are 3 lakes in the park, the tree of them are written "G1", "G2" and "G3".

We are performing MCA with the package FactoMineR.

```{r, include=FALSE}
library(FactoMineR)
res.mca = MCA(hippo, quanti.sup=1)
```

Let's have a look to Eigen Values:

```{r}
res.mca$eig
```

The total variance of the data set is divided between four dimensions. The first dimension concentrate 32.6% of the total variance. The first plan gather 58% of the variance.

Here, you can see the results for the individuals, their coordinates, contribution and the quality of their projection.

```{r}
head(res.mca$ind$coord)
head(res.mca$ind$cos2)
head(res.mca$ind$contrib)
```

Coordonates of the individuals are their position on the first plan. Cos2 represents the quality of the representation of the individuals on the first plan. Contribution is how the point contribute to the creation of different axis.

```{r}
res.mca$call$Xtot

res.mca$var$coord
res.mca$var$contrib
res.mca$var$cos2

res.mca$quanti.sup$coord
```

Let's see how to interprete these graphs:

The two axis

## Take Home Message

MCA is a statistical method adapted to table of type "individuals x quatitative variable". Eingen Values correspond to means of squered correlation ratios. It could be used a pre-processing before a classification or a coinertia analysis on tables with quantitative.

# PART 3: RDA

## Mathématiques

## Interprétation

This chapter is a simple example using R

You can import R package using the code

```{r}
library(tidyverse)
```

and then describe the purpose of your chapter as well as executing R command.

For example a basic summary of a dataset is given by

```{r}
df <- read.table("https://gist.githubusercontent.com/slopp/ce3b90b9168f2f921784de84fa445651/raw/4ecf3041f0ed4913e7c230758733948bc561f434/penguins.csv", sep = "," , header = TRUE)
```

and produce a graph

```{r}
df %>% ggplot() +
	aes(x=species, y = body_mass_g) +
	geom_boxplot()  
```

A citation @bauer2023writing

# PART 4: RLQ

## Mathématiques

## Interprétation

This chapter is a simple example using R

You can import R package using the code

```{r}
library(tidyverse)
```

and then describe the purpose of your chapter as well as executing R command.

For example a basic summary of a dataset is given by

```{r}
df <- read.table("https://gist.githubusercontent.com/slopp/ce3b90b9168f2f921784de84fa445651/raw/4ecf3041f0ed4913e7c230758733948bc561f434/penguins.csv", sep = "," , header = TRUE)
```

and produce a graph

```{r}
df %>% ggplot() +
	aes(x=species, y = body_mass_g) +
	geom_boxplot()  
```

A citation @bauer2023writing
