---
title: "Introduction to bayesian modeles for ecology"
author : Coline - Jules - Mathis - Romane - Laura
bibliography: references.bib
execute: 
  freeze: auto
output: 
  html_document:
   toc: true
   toc_float: true
---

```{r setup}
#| include: false

library(tidyverse)
library(deSolve)
```

# Introduction

-   History of Bayesian concept.
-   Philosophy behind it. Frequentist approach : "What is the probability to observe my data given my model."

Bayesian approach : "What is the probability of my model given my data."

# Genrale theory behind bayesian approach.

Explanation in the special case of conjugation between prior and posterior

## Main objectif, for context
Our goal is to understand in the most simple way how Bayesian statistic estimate parameters of natural process. To do so, we will use simple dynamical models to produce data. With those models every parameters is known, we will try to find those parameters with the observations.

## An exemple : estimate one varaibles
The most simple model : Malthusian growth model

```{r}
expo.mod <- function(t,N,r){list(N*r)}
N0 = c(y=1)
r = 0.1
dyna = ode(y = N0, times = 1:100, func = expo.mod, parms = r )
dyna = data.frame(dyna)
 
ggplot(dyna)+
  geom_line(aes(time,y))
```

```{r}

```



## Some math






# When aproximation by computation is needed

## Aproximation with MCMC