---
title: "Ordinary Differential Equations and their application in Ecology"
bibliography: references.bib
execute: 
  freeze: auto
output: 
  html_document:
   toc: true
   toc_float: true
---

Introduction to ODE's and their applications in Ecology

Dans cours de Cédric:
I] Les systèmes dynamiques (notion de chronologie, inconnue, variable, etc.) + Exemple d'un système très simple avec naissances ; migration ; mortalité ;

II] Définition d'une équation différentielle. + un peu d'histoire, ce qui permet de l'intégrer à l'écologie à partir du XXiè siècle.

III] Les différents modèles 
  * Modèles discrets (Equation aux différences ?)
  * Modèles continus (EDOs ?)
  
Modèle de MAlthus
Modèle de Verhulst

IV] Les analyses possibles




This chapter is a simple example using R

You can import R package using the code

```{r}
library(tidyverse)
```


and then describe the purpose of your chapter as well as executing R command.


For example a basic summary of a dataset is given by 

```{r}
df <- read.table("https://gist.githubusercontent.com/slopp/ce3b90b9168f2f921784de84fa445651/raw/4ecf3041f0ed4913e7c230758733948bc561f434/penguins.csv", sep = "," , header = TRUE)
```

and produce a graph

```{r}
df %>% ggplot() +
	aes(x=species, y = body_mass_g) +
	geom_boxplot()  
```


A citation @bauer2023writing
