title: "Ordinary Differential Equations and their application in Ecology" author: "Amélie Cocchiara, Théo Laguilliez, Clément Monaury, Laura Martinez Anton, Anastasia Paupe" date: "`r Sys.Date()`" bibliography: references.bib execute: freeze: auto output: html_document: toc: true toc_float: true editor: markdown: wrap: 72

```{r}
#Add packages
  df_print: paged


<style>
body {
text-align: justify}
</style>
```

```{r init, include=FALSE}
# Configure your session and load R packages
rm(list=ls()) # Properly clear workspace

# Install packages
pkg <- c("knitr", "ggplot2", "tidyverse", "draw", "devtools")
inst.pkg <- row.names(installed.packages())
pkg2inst <- pmatch(pkg, inst.pkg)
if (any(is.na(pkg2inst[1:4]))) install.packages(pkg[which(is.na(pkg2inst[1:4]))],
                                                repos = "http://cran.at.r-project.org")
# Load packages
library(knitr)
opts_chunk$set(echo = FALSE, comment = "", cache = TRUE, fig.align = "center")
library(ggplot2) # graph package
library(tidyverse)
library(draw) # To draw some shapes
library(devtools)
devtools::install_github("hadley/emo")
```

```{r, include = FALSE}
# Generation of the needed emoji
emo::ji("alien")
emo::ji("goat")
emo::ji("vomit")
emo::ji("silly")
emo::ji("sunglasses")
emo::ji("dork")
```

# Introduction & objectives

Welcome to you, young Master 1 student!

It's now been ?x?(= number of chapters before ours) chapters since you plunged into this adventure full of numbers, formulas and other mathematical oddities (🤢). At the moment, you're probably feeling a strong need for a change of orientation. I mean, we're doing ecology, not maths! And raising goats in Auvergne sounds pretty cool...kinds of...🐐

But don't give up! The first (really interesting 😜) chapter is right there in front of you!

Are you interested in population and epidemiological trends and dynamics? Would you like to be able to predict the future (and the past, no kidding) from the present?

Then this is the chapter for you! Work it as much as you can! You'll end up shining in modelling (and maybe even in society).

At the end, you'll be able to ...

# I. Dynamic systems

Let's start simply!

First of all, let's define the framework of the study (= cadre de l'étude).

We consider a dynamic system. Or in other words, a physical system whose state changes as a function of time. By "state", we mean "physical variables" such as a number of individuals, a position, a temperature, etc., which are functions of time (= which evolve over time).

For example, suppose you wanted to study the evolution over time of a population of aliens newly landed on Earth (yes, it's possible, I saw that in men in black 😎).

[Your system is]{.underline}: the Alien population.\
[The state of this population is]{.underline}: the number of individuals, noted '👽'.\
This state is a function of time (the number of individuals changes over time). We therefore note it: '👽(t)'.

So your aim is to study this system. But as you can imagine, you can't start from scratch! To study the evolution of such a system, you need to know:\
- **its initial state**: values of its characteristic variables at the initial time of the study (t=0).\
- **the processes underway in the biological system**: the way of gaining (i.e. birth, migration) and/or losing (i.e. mortality) individuals for example. Don't hesitate to draw the system on a piece of paper, to avoid getting lost (*Figure 1*)!\
- **the parameters associated with these processes**

Let's make it more concrete:

```{r, include=TRUE, fig.height=2.8, fig.width=5}
# Create a new drawing page and set the drawing settings
drawPage(width = 5, height = 5, units = "inches")

# Draw the different shapes of the system on the page
#     * The rectangle
drawBox(x = 2.5, y = 2.5, width = 1, height = 1)

#     * The mortality arrow 
drawLine(x = c(3, 4),
         y = c(2.5, 2.5))
drawLine(x = c(3.9, 4),
         y = c(2.4, 2.5))
drawLine(x = c(3.9, 4),
         y = c(2.6, 2.5))

#     * The migration arrow 
drawLine(x = c(1, 2),
         y = c(2.5, 2.5))
drawLine(x = c(1.9, 2),
         y = c(2.4, 2.5))
drawLine(x = c(1.9, 2),
         y = c(2.6, 2.5))

#     * The birth 'arrow' (its a circle) 
drawCircle(x = 2.5,
           y = 3.4,
           radius = 0.4)
drawLine(x = c(2.5, 2.6),
         y = c(3, 3.1))
drawLine(x = c(2.5, 2.6),
         y = c(3, 2.9))

# Add the text
#     * For the population
drawText(x = 2.5, y = 2.5, text = "Population")
drawText(x = 2.5, y = 2.3, text = "👽(t)")

#     * For the mortality
drawText(x = 4.45, y = 2.5, text = "Mortality")
drawText(x = 3.5, y = 2.35, text = "d 👽(t)")

#     * For the migration
drawText(x = 0.5, y = 2.5, text = "Migration")
drawText(x = 1.4, y = 2.35, text = "m")

#     * For the birth
drawText(x = 2.5, y = 3.4, text = "Birth")
drawText(x = 3.2, y = 3.4, text = "b 👽(t)")

# Add the caption
drawText(x = 2.5, y = 1.5, text = "Figure 1: Diagram of the 'Alien population' dynamic system")

```

To keep things simple, we'll take a discrete-time simulation, with a time step of 1.

Let define t like that: $$t \in \mathbb{N}^+$$ This means that t belongs to the set of natural integer numbers. t can therefore take the values 0, 1, 2, 3, 4, 5, etc..

Here, the variable of interest (= the state of the system) is the number of individuals in the population. This number changes over time, and is therefore denoted 👽(t). The biological processes responsible for this change over time are :\
- **mortality**: noted d (for death). d👽 corresponds to the proportion of individuals that will die in 1 unit of time\
- **birth**: denoted b. This means that 👽 individuals will produce b👽 new individuals in 1 unit of time\
- **migration**: denoted m. This means that 👽 increases by m individuals in 1 unit of time.

```{r, include=TRUE, fig.height=5, fig.width=6}
# Create a new drawing page and set the drawing settings
drawPage(width = 5, height = 4, units = "inches")

# Draw the orthogonal plan on the page
#     * The x axe 
drawLine(x = c(0.5, 4),
         y = c(0.5, 0.5),
         lineWidth = 2.5)
drawLine(x = c(3.9, 4),
         y = c(0.4, 0.5),
         lineWidth = 2.5)
drawLine(x = c(3.9, 4),
         y = c(0.6, 0.5),
         lineWidth = 2.5)
#     * The y axe 
drawLine(x = c(0.5, 0.5),
         y = c(0.5, 4),
         lineWidth = 2.5)
drawLine(x = c(0.4, 0.5),
         y = c(3.9, 4),
         lineWidth = 2.5)
drawLine(x = c(0.6, 0.5),
         y = c(3.9, 4),
         lineWidth = 2.5)

# Draw the 'discret' time lines
#     * The vertical one 
drawLine(x = c(2, 2),
         y = c(0.5, 2),
         lineWidth = 1)
#     * The horizontal one
drawLine(x = c(0.5, 2),
         y = c(2, 2),
         lineWidth = 1)

# Draw the 'continuous' time lines
#     * The vertical one 
drawLine(x = c(3, 3),
         y = c(0.5, 3),
         lineWidth = 1)
#     * The horizontal one 
drawLine(x = c(0.5, 3),
         y = c(3, 3),
         lineWidth = 1)

# Draw of the lines
drawCurve(x = c(1.5, 4.1),
          y = c(1, 3.5),
          angle = 80, # Angle of the curve
          open = T, # No link btw starting and ending points
          inflect = F, # No infection points
          curvature = -0.5,
          points = 5,
          square= F,
          squareShape = 0,
          lineWidth = 1.5)
drawLine(x = c(1, 3.7),
         y = c(1, 3.7),
         lineWidth = 1.5,
         lineColor = "blue")
drawLine(x = c(1, 3.7),
         y = c(1, 3.7),
         lineWidth = 1.5,
         lineColor = "blue")

# Add the text
drawText(x = 0.1, y = 4, text = "Number")
drawText(x = 0, y = 3.8, text = "of individuals")
drawText(x = 4.2, y = 0.4, text = "Time")
drawText(x = 3, y = 0.4, text = "(t+Δt)")
drawText(x = 2, y = 0.4, text = "(t)")
drawText(x = 0.1, y = 3, text = "👽(t+Δt)")
drawText(x = 0.1, y = 2, text = "👽(t)")



# Add the caption
drawText(x = 2.80, y = 0, text = "Figure 2: Schematic graphic representation of the growth")
drawText(x = 3.15, y = -0.2, text = "of a population over time")

```

$$
\begin{align} & NPZ = \begin{cases} \frac{dP}{dt}= \frac{V_m N}{k_s + N}P -mP -ZR_m(1-e^{-\Lambda P}) & \text{Pythoplankton}
\\ \frac{dZ}{dt}= \gamma ZR_m(1-e^{- \Lambda P}-dZ) & \text{Zooplakton}
\\ \frac{dN}{dt}=-\frac{V_m N}{k_s + N}P+mP+dZ+(1- \gamma ) ZR_m (1-e^{- \Lambda P}) & \text{Nutrients} \end{cases}
\\ 
\\ & \text{Definitions of variable:}
\\ & - V_m: \ \text{Maximum phytoplankton growth rate}
\\ & - N: \ \text{Nutrient concentration}
\\ & - k_s: \ \text{Half saturation constant for nutrients}
\\ & - P: \ \text{Phytoplankton stock size}
\\ & -m: \ \text{Phytoplankton mortality rate}
\\ & -Z: \ \text{Zooplankton stock size}
\\ & - \gamma : \ \text{Zooplankton growth efficiency}
\\ & - R_m : \ \text{Maximum zooplankton ration}
\\ & - \Lambda : \ \text{Ivlev constant}
\\ & -d: \ \text{mortality rate} \end{align} $$

$$\Delta$$ $$\int_a^b 👽(t) \, \mathrm dt$$

# II. Order 1 ODE & definition of ODE

# III. Order 2 ODE

# **IV. 3-equation system**

We will now make the system more complex by adding an additional equation. To solve these systems, two methods are possible. If we are in the case of a closed population then we can remove an equation: see example SIR model. If we are in the case of an open population: see example of the Lotka-Volterra model with 3 species.

## 1) SIR model and resolution by hand

The most used **models in epidemiology** are mathematical models of infectious diseases such as the SIR model. These models are based on compartmentalization that divide the population into various possible disease states. The **SIR model** is composed of 3 compartments: initially healthy or **susceptible individuals** (S), **infected individuals** (I) and cured individuals or **resistent** (R).

$$SIR=\begin{cases}\frac{d_S}{d_t}=- \alpha SI+\gamma R &\text{Susceptible} \\ \frac{d_I}{d_t}=\alpha SI-\beta I&\text{Infected} \\ \frac{d_R}{d_t}=\beta I- \gamma R &\text{Resistent}\end{cases} 
\\
\\ \begin{align} With : & - \alpha \text{: disease transmission rate}
\\ & - \beta \text{: cure rate}
\\ & - \gamma \text{: coefficient of loss of immunity} \end{align}$$

We assume a **closed population**, so : $$ S+I+R = 1 \Leftrightarrow R=1-S-I $$

Thus, we **can express R from S and I**, so we can remove the equation from our system and we end up with a system with 2 equations.

### Balance points

We look for balance points such as:

$$ SIR=\begin{cases}\frac{d_S}{d_t}=0 &\text{Susceptible} \\ \frac{d_I}{d_t}=0&\text{Infected} \end{cases} $$

To do this, we solve :

$$ \begin{align} & \begin{cases} - \alpha SI-\gamma R = 0 &\text{Susceptible} \\ \alpha SI-\beta I =0 &\text{Infected} \end{cases}
\\ \Leftrightarrow & \begin{cases} \alpha SI-\gamma (1-S-I) = 0 \\ \alpha SI-\beta I =0 \end{cases} \end{align} $$

We first focus on solving:

$$\begin{align} & \alpha SI - \beta I=0
\\ \Leftrightarrow \ & I(\alpha S- \beta) =0 \\ \Leftrightarrow \ & I=0\ or\ S= \frac{\beta}{\alpha} \end{align}$$

**If *I = 0***, we integrate I=0 into our "susceptible" equation to obtain the solution for S:

$$\begin{align} & \ \ \ \ \ \begin{cases} \alpha S *0 + \gamma (1-S-0)=0 \\ I=0 \end{cases} \\ & \Leftrightarrow \begin{cases} \gamma (1-S)=0 \\ I=O \end{cases} \\ &\text{We thus obtain our first solution:} \\ &\Leftrightarrow \begin{cases} S=1 \\ I=0 \end{cases} \end{align} $$

It's the **disease-free equilibrium** (DFE).

**If *S = b/a***, we integrate S into our "susceptible" equation to obtain the solution for I:

$$
\begin{align} & \ \ \ \ \  \begin{cases} - \alpha \frac{\beta}{\alpha} I + \gamma (1- \frac{\beta}{\alpha} - I) =0
\\ S = \frac{\beta}{\alpha} \end{cases}
\\ & \Leftrightarrow \begin{cases} -\beta I + \gamma - \frac{\beta \gamma}{\alpha} -\gamma I = 0 \\ S = \frac{\beta}{\alpha} \end{cases}
\\ & \Leftrightarrow \begin{cases} \beta I + \gamma I = \gamma (1 - \frac{\beta}{\alpha}) \\ S= \frac{\beta}{\alpha} \end{cases}
\\ & \text{We thus obtain our first solution:}
\\ & \Leftrightarrow \begin{cases} I= \frac{\gamma (1- \frac{\beta}{\alpha})}{\beta + \gamma} \\ S=\frac{\beta}{\alpha} \end{cases} \end{align} 
$$

It's the **endemic equilibrium** (EE).

### Jacobian matrix

Now, let's perform linear stability analysis for each equilibrium point by computing the Jacobian matrix :

$$ J=\begin{bmatrix} \frac{\delta S}{\delta S} & \frac{\delta S}{\delta I} \\ \frac{\delta I}{\delta S} & \frac{\delta I}{\delta I}  \end{bmatrix} \Leftrightarrow \begin{bmatrix} - \alpha I- \gamma & - \alpha S- \gamma \\ \alpha I & \alpha S- \beta \end{bmatrix} $$

Analyze the stability of each equilibrium point by examining the signs of the real parts of the eigenvalues. If the real parts are all negative, the equilibrium is stable; if any are positive, the equilibrium is unstable. The stability will depend on the specific values of the parameters **α** and **β**.

### For the disease-free equilibrium :

$$ J(1,0)= \begin{bmatrix} -\gamma & - \alpha -\gamma \\ 0 & \alpha - \beta \end{bmatrix} $$

The Jacobian matrix for the pair (1,0) is an upper triangular matrix, so the eigenvalues are the diagonal elements : $$ \lambda _1 = - \gamma \ \text{and} \ \lambda _2 = \alpha - \beta $$

γ is a coefficient representing loss of immunity, so γ\>0.

So : $$ \begin{align} & \lambda _1 <0 
\\ & \text{If:} \ - \alpha > \beta , \ \lambda _2 > 0 \ \text{Unstable equilibrium point}
\\& \ \ \ \  - \alpha < \beta , \ \lambda _2 < 0 \ \text{Asymptotically stable point}  \end{align} $$

In order to better understand the behavior of the unstable point, we look for the sign of the determinant of the matrix:

$$ det(J(1,0))= - \gamma (\alpha - \beta ) \\
\Leftrightarrow \ \gamma (\beta - \alpha)
\\ \gamma > 0 \ and \ \alpha > \beta
\\ det(J(1,0)) < 0 \Rightarrow \text{Saddle point} $$

Saddle points exhibit a combination of stable and unstable behavior in different directions in the state space.

### For the endemic equilibrium

$$ J(\frac{\beta}{\alpha},\frac{\gamma (1-\frac{\beta}{\alpha})}{\beta + \gamma})= \begin{bmatrix} \frac{\gamma (\beta- \alpha)}{\beta + \gamma} - \gamma & - \beta -\gamma \\ \frac{\gamma (\alpha - \beta)}{\beta + \gamma} & 0 \end{bmatrix} $$

Here, we do not have a characteristic matrix so to determine the eigenvalues, it will be necessary to solve the characteristic polynomial. We will first look at the sign of the determinant of the matrix to study the behavior of the system.

So, we calculate the determinant : $$\begin{align} & det(J(\frac{\beta}{\alpha},\frac{\gamma (1-\frac{\beta}{\alpha})}{\beta + \gamma})) = - (- \beta - \gamma) * \frac{\gamma (\alpha - \beta)}{\beta + \gamma} = \gamma (\alpha - \beta) \end{align}$$

If **α** \< **β**, det(J) \< 0 : a saddle point.

If **α** \> **β**, det(J) \> 0 : look the trace of the matrix.

$$Tr(J(S,I)) = \frac{\gamma (\beta - \alpha)}{\beta - \gamma} - \gamma
\\
\\ \gamma > 0 \ \text{and} \ \alpha > \beta
\\
\\ \text{So:} \ Tr(J(S,I)) <0 $$

In this case we have a stable point. The specific type of stability depends on the signs and nature of the eigenvalues.

Now, we solve the characteristic polynomial to obtain the eigenvalues:

$$ det(J- \lambda Id) = 0
\\ \text{with Id : identity matrix}
\\
\\ \Leftrightarrow \begin{bmatrix} \frac{\gamma (\beta - \alpha)}{\beta + \gamma}-\gamma -\lambda 
& - \beta - \gamma
\\  \frac{\gamma (\alpha - \beta)}{\beta + \gamma} & - \lambda \end{bmatrix} = 0
\\ \Leftrightarrow - \lambda (
\frac{\gamma (\beta - \alpha)}{\beta + \gamma}-\gamma -\lambda) 
- (- \beta - \gamma )(
\frac{\gamma (\alpha - \beta)}{\beta + \gamma}
) = 0 
\\ \Leftrightarrow \lambda^2 + \lambda (\gamma +  \frac{\gamma (\alpha- \beta)}{\beta + \gamma})+ \gamma (\alpha - \beta) = 0 $$

The number of solutions is indicated by the value of the discriminant : $$ \Delta = (\gamma +  \frac{\gamma (\alpha - \beta)}{\beta + \gamma})^2 - 4 (\gamma (\alpha - \beta))
\\ \text{With} \ \alpha > \beta
\\
\\
\begin {align} \text{If} \ & \Delta < 0:\ \text{you will have two complex conjugate roots.} \\ &\Delta = 0:\ \text{you will have one real repeated root.} \\ &\Delta>0: \ \text{you will have two distinct real roots.} \end{align} $$

From now,we will assign **numerical values**: $$ \alpha = 4, \ \beta = 2 \ and \ \gamma = 1 
\\
\Delta = (1 + \frac{1(4-2)}{1+2})^2 - 4(1(4-2))= -5.22 
\\ So \ \Delta < 0 $$

$$ \text{Solutions are given by the following formula:}
\\ \lambda= \frac{-(\gamma + \frac{
\gamma (\alpha - \beta)}{\beta + \gamma}) \pm \sqrt \Delta}{2}i
\\ \frac{\sqrt \Delta}{2}i: \ \text{imaginary part.}
\\ \frac{-(\gamma + \frac{\gamma (\alpha - \beta)}{\beta +\gamma})}{2}: \ \text{real part.} 
\\
\\ \text{Real part}= -\frac{5}{3}$$

The stability of a linear system is often determined by the real parts of the eigenvalues. In this case, the real part is negative which indicates a **stable attractive node**.

This sounds complicated, **but don't panic** ! In ecology we are lucky to have a great tool, R, which allows us to resolve this type of system. We will see how can R help us .

### Solving SIR model with R

Use R to study the behavior of our model as double interest.

-   Verify if our theoretical calcul is okay

-   If you can't do your calcul help you to have an idea of the behavior of the model

To show how you can use phaseR and deSolve to solve ODE we will use the SIR we just solved. We hope our calcul is good. In this chapter, we will show what's the most common function you can use, but not all the parameters you can add. I know you are probably sobbing "wHyYyYyy ?" (Who has never dreamed of knowing all the difference between the Euler and Runge Kunta order 4 ?) but we let you make your search if you need more information.

#### Step 1: Prepare you environnement

Let's start by erasing all the data in our environment in order to don't bother us when we work. Then we load the 3 packages we will use.

-   The most famous library to solve and study ODE is deSolve you can also compute yourself if you want but it will be a harsh task. A great point of deSolve it's its simplicity to use and we hope you will agree with us a the end of this chapter

-   phaseR Performs a qualitative analysis of one- and two-dimensional autonomous ordinary differential equation systems, using phase plane methods

-   We will also use ggplot to make a beautiful graph. It's also a little more intuitive way to create and manipulate plot than using a basic ploting system in r

-   rgl is a package which give you tools to creat 3d plot and objects

-   tidyr is like ggplot but for data manipulation. It doesn't bring very special tools but able us to transform easily our table

As we talk about programmation we have to talk about technical issues you can have. To prevent you from committing this we put during the presentation some technical issues to think of before coding. If you want to do the same things as us you will probably find yourself in some problems don't be afraid R is widely used you will easily find people who have defeated the same problem. Don't be ashamed to search for help!

```{r message=F, warning=FALSE}
#We erase all data use before
rm(list = ls())
#load library we need
library(ggplot2)
library(deSolve)
library(phaseR)
library(rgl)
library(tidyr)
```

#### Step 2: Define your model

It's time to start modeling! The first step is to create our function which gives us the evolution of our population according to time. This function takes a step of time, the initial population value, and the value of different parameters. The output of this function needs to be a list of the new populations. To have the behavior of our function we have to use the same hypothesis as in the calculation part so we will not code how the Recover compartment evaluates (R). If we need the Recover evolution we will use: 
$$1 = S + I + R \Leftrightarrow R = 1 - S - I $$
Technical consideration: When you build your function you are free to make some modification BUT the order of the input has to be time, population, and parameter value because the ODE function want you to do this!

```{r}

mod_SIR <- function(t , dy , parameters){
#we take our population and give give to to diffirent variable
  S <- dy[1]
  I <- dy[2]

#We give the value of parameters to the variable (there is greater way to do this but i let you find out)
  alpha <- parameters[1]
  beta <- parameters[2]
  gamma <- parameters[3]

#sol is the vector where we stock our solution
  sol <- numeric(2)

#We compute the equation need
  sol[1] <- - alpha * S * I + gamma * (1 - S - I)
  sol[2] <- alpha * S * I  - beta * I
  
#And list solution 
  list(sol)
}

```

#### Step 3: Study how the system evaluates among time

Now we have done the heart of our model let's study it with ODE. It's from the deSolve package to solve our model. To do this we define some quantities. For the time we use the seq function which creates a vector that goes from 0 to 600 by 0.01 step. For parameter values, we create a vector in the same order as in our function. And value we start to form

Then we use ODE to simulate how our population evolves at each time in our vectors. We take the output from ODE and manipulate it a little bit to make visualization easier.

```{r}
#Definition of parameters we will need for the suite
param <- c(alpha = 0.2, beta = 0.1, gamma = 0.2)
times = seq(from = 0 , to = 600, by = 0.01)
depart = c(0.99 , 0.001 )

#Calcul of the evolution of our population 
resolnum <- ode(y = depart, times = times, func = mod_SIR , parms = param)

#Transfortmation of data in order to be use easely
resolnum <- as.data.frame(resolnum)
resolnum$R<- 1-  resolnum[,2] + resolnum[,3] 

colnames(resolnum) <- as.factor(c("Times" , "Susceptible" ,"Infectious" ,"Recovered"))



#Plotting time
ggplot(resolnum,aes(x=Times))+
  geom_line(aes(x = Times, y = Susceptible),color ="chartreuse" ,size =1.5)+
  geom_line(aes(x = Times,y = Infectious),color = "red" ,size =1.5)+
  geom_line(aes(x = Times,y = Recovered),color = "darkviolet" ,size =1.5)+
  labs(title = "Population evolution among times", y = "Population")+
  theme_classic()


```

Now we have an idea of how our population evaluates thought time we can try to find the behavior

#### Step 4: Equilibrium point and behavior

Let's study the behavior of our model and its equilibrium points. We're going to study S as a function of I First function displays model behavior Second draws the lines where the models cancel out.

Technical consideration: To have all nullclines you should extend the area you want to explore. For example, we only have interest in the area where S and I go to 0 at 1(we made the hypothesis the sum of all compartments should be equal to 1) but in our simulation, we go from -0.1 to 1.1.

```{r}
#Plot the velocity field
  flowField(mod_SIR ,xlim=c(-0.1,1.1),ylim=c(-0.1,1.1),parameters = param,add = F,xlab = "S" , ylab = "I")
  #Draw the isoclines of our model
nullclines(mod_SIR ,xlim=c(-0.1,1.1),ylim=c(-0.1,1.1),parameters = param,points = 100, add = T,add.legend = F,size =2)

```

As you can see we have two equilibrum points. The first is at the center of our phase diagram, and the arrows seem to show that we're tending towards this point. A second is present where I = 0 and S = 1 on the contrary, we seem to be moving away from it.
If this graph allows us to observe 2 points of equilibrium, it is not easy to determine their behavior.

To look at the behavior near these points, we can use the trajectory function, which will plot the trajectories of our system as a function of a starting point.

```{r message = FALSE}


#we define our initials conditions
X0 <- sample(x = seq(0,1,0.01),size =1)
Y0 <- sample(x = seq(0,1-X0,0.01),size =1)
cond_init = c(X0, Y0)
#We simulate what's happened for one trajectory
  trajectory(mod_SIR ,cond_init,tlim= c(0,50),parameters = param,col = "black",add = F,xlim=c(0,1),ylim = c(0,1))
  
#We repeat the same process for 50 other trajectory
for(i in 1:50){
  X0 <- sample(x = seq(0,1,0.01),size =1)
Y0 <- sample(x = seq(0,1-X0,0.01),size =1)

cond_init = c(X0, Y0)
  trajectory(mod_SIR ,cond_init,tlim= c(0,50),parameters = param,col = i,add = T)
}
```
What the vector field showed us seems to be confirmed by this graph. Let's check our observations with a new functions (promise they're the last)


We'll check the behavior of the center point using the stability function. We could also have used the findEquilibrum function to find all the equilibrium points, but we'll let you have your fun.

```{r}
# Checking stability of the point we see
stability(SIR,ystar = c(0.5,0.5),parameters = param )$classification
```
The two functions confirm our observations and the calculations made just before - what more could you ask for?

There are still other functions in the deSolve and pahse R packages that could be of interest, but which are useful in more specific cases, so we won't look at them here.

## 2) Lotka-Volterra : A three-species model

Biotic interactions are complex, requiring the interaction of numerous parameters. If we take the **Lotka-Volterra** example, it only considers the **interactions between two species**, prey and predator. It assumes a constant mortality rate among predators, regardless of prey. However, **predators are themselves the prey of other super-predators**. Their mortality rate should therefore depend on the super-predators they encounter. Thus, the **prey-predator-superpredator** extension of the Lotka-Volterra model would be a good representation of natural trophic interactions.

In this case, we considerer a **superpredator**, a **predator** an a **prey**.

A superpredator is an organism that reaches adulthood at the top of the food chain. Thus, by definition, a superpredator is not the prey of any other species. The predator is consider to be in the middle of the food chain and the prey at the bottom.

The model can be written as the following system of differential equations:

$$ PPS=\begin{cases} \frac{d_x}{d_t}=ax-bxy &\text{Prey} \\ \frac{d_y}{d_t}=-cy+dxy-eyz&\text{Predator} \\ \frac{d_z}{d_t}=-fz+gyz &\text{Superpredator} \end{cases} $$

The model parameters are :

-   *a* : *x* natural growth rate

-   *-f* and *-c* : natural death rate of *y* and *z*

-   The growth rate of *y* and *z* depends on the number of prey caught. It is modeled by *dxy* for *y* and by *gyz* for *z*.

-   The death rate of *x* and *y* depends on the number of top predators *y* for *x* and *z* for *y* . It is modeled by *-bxy* for *x* and by *-eyz* for *y*.

This model is only useful for chains of predators. It doesn't take into account the presence of several predators with different dynamics on the same prey.

Let's try to solve this system: To do this, the first step is to find the balance of the model, that is to say the values (x,y,z) so that:

$$\frac{d_x}{d_t}=\frac{d_y}{d_t}=\frac{d_z}{d_t}=0$$

We take our previous systems that we factored:

$$ PPS=\begin{cases}\frac{d_x}{d_t}=x(a-by) &\text{Prey} \\ \frac{d_y}{d_t}=y(dx-c-ez)&\text{Predator} \\ \frac{d_z}{d_t}=z(gy-f) &\text{Superpredator}\end{cases} $$

The obvious solution is (0,0,0) but is not the only one. Indeed, for the derivatives to be equal to 0, then we must find the following solution:

$$(a-by)=(gy-f)=(dx-c-ez)=0$$

We first solve : $$a-by=0\Leftrightarrow a=by \Leftrightarrow y=\frac{a}{b}$$ $$gy-f=0\Leftrightarrow f=by \Leftrightarrow y=\frac{f}{g}$$

So, a solution for a yeq is: $$y_{eq}=\frac{a}{b}=\frac{f}{g}$$ Therefore, there is an equilibrium when : $$ga=fb$$

In order to determine the other solutions, we must solve the following jacobian matrix : $$ J=\begin{bmatrix}a-by&-bx&0\\dy&-c+dx-ez&-ey\\0&gz&gy-f\end{bmatrix}$$

The steps are the same as for the SIR model but here the matrix is 3x3. We invite you to look at the chapter on matrix calculation to determine the solutions.

Finding equilibrum point and their behaviour is a little bit harder than before so it's time to use R.

#### Solving Lotka-Volterra with R

Hey R is back! You love last part with how to solve 2 equation ODE in R ? Now let's find out how to solve a 3 equation system!
Unfortunately, trying to solve this equation by hand leads to problems: it's not easy to find all the equilibrium points. As mentioned in the previous section, phaseR is not suitable for solving three-equation systems.
We can compute our function but unless you a nerd or MODE you don't want to do this. the solution we choose in this paragrph it's to only study numercally system.

So let's start again what we just do! We use the same library

```{r}
rm(list = ls())
library(ggplot2)
library(deSolve)
library(phaseR)
library(tidyverse)
library(rgl)
```

We create a function with the same limitation as in our first example     .

```{r}
PPS <- function(t , dy , parameters){

  x <- dy[1]
  y <- dy[2]
  z <- dy[3]
  
  r1 <- parameters[1]
  r2 <- parameters[2]
  r3 <- parameters[3]
  b1 <- parameters[4]
  b2 <- parameters[5]
  c1 <- parameters[6]
  c2 <- parameters[7]
  
  sol <- numeric(3)
  
  sol[1] <- r1 * x - b1 * y * x
  sol[2] <- - r2 * y + b2 * x * y  - c1 * y * z 
  sol[3] <- - r3 * z + c2 * z * y 
 
  list(sol)
}

```

In order to determine the behavior of our system we should use the ode function to make many graph and try to understand numerically which parameter ar important. In the following example we pass this step and present you 3 cases we can have in our system

```{r}
#Paramters definition
param <- c(r1 = 0.5, r2 = 0.2 , r3 = 0.1, b1 = 0.04, b2 = 0.04, c1 = 0.09, c2 = 0.01)
temps = seq(from = 0 , to = 100, by = 0.01)
depart = c(20 , 10 , 5)
#Simulation
resolnum <- ode(y = depart, times = temps, func = PPS, parms = param)
resolnum <- as.data.frame(resolnum)
colnames(resolnum) <- c("Times" , "Prey" ,"Predators" ,"SuperPredators")

#Plotting some results


ggplot(resolnum)+
  geom_line(aes(x = Times, y = Prey),color ="chartreuse" ,size =1.5)+
  geom_line(aes(x = Times,y = Predators),color = "red" ,size =1.5)+
  geom_line(aes(x = Times,y = SuperPredators),color = "darkviolet",size =1.5)+
  theme_classic()


ggplot(resolnum)+
   geom_point(aes(x = Prey, y = Predators),color ="black" ,size =1.5)+
  theme_classic()

ggplot(resolnum)+
   geom_point(aes(x = Prey, y = SuperPredators),color ="black" ,size =0.2)+
  theme_classic()

ggplot(resolnum)+
   geom_point(aes(x =  SuperPredators, y = Predators),color ="black" ,size =1.5)
+theme_classic()

plot3d(resolnum$Prey,resolnum$Predators,resolnum$SuperPredators,xlab ="Prey",ylab ="Predators",zlab ="SuperPredators",lit =F)
```
This first case is a very special one, since we observe a runaway system, with the number of prey increasing exponentially without being limited by predators. While mathematically this model makes sense, it is not very common in nature, due to the absence of infinite resources.



```{r}
param <- c(r1 = 0.01, r2 = 0.2 , r3 = 0.1, b1 = 0.04, b2 = 0.04, c1 = 0.09, c2 = 0.01)
temps = seq(from = 0 , to = 1000, by = 0.01)
depart = c(20 , 10 , 5)
resolnum <- ode(y = depart, times = temps, func = PPS, parms = param)

resolnum <- as.data.frame(resolnum)
colnames(resolnum) <- c("Times" , "Prey" ,"Predators" ,"SuperPredators")



ggplot(resolnum)+
  geom_line(aes(x = Times, y = Prey),color ="chartreuse" ,size =1.5)+
  geom_line(aes(x = Times,y = Predators),color = "red" ,size =1.5)+
  geom_line(aes(x = Times,y = SuperPredators),color = "darkviolet",size =1.5)+
  theme_classic()

ggplot(resolnum)+
   geom_point(aes(x = Prey, y = Predators),color ="black" ,size =1.5)
+theme_classic()

ggplot(resolnum)+
   geom_point(aes(x = Prey, y = SuperPredators),color ="black" ,size =0.2)
+theme_classic()

ggplot(resolnum)+
   geom_point(aes(x =  SuperPredators, y = Predators),color ="black" ,size =1.5)
+theme_classic()

plot3d(resolnum$Prey,resolnum$Predators,resolnum$SuperPredators,xlab ="Prey",ylab ="Predators",zlab ="SuperPredators")
```

In this case, our system moves towards a balance between prey and predators. This time, however, superpredators are excluded from our system.

For a more thorough analysis, we'd have to simulate the same model but draw our parameters randomly, but unfortunately this takes a lot of time, so we'd have to do it with very few simulations.
To get a better idea of the model's behavior, more advanced simulations should be carried out, but unfortunately the fun has to stop at some point.

If you eat graphics cards and want to delve deeper into the subject, the book Solving Differential Equations in R may be of some interest, even if it deals with much more complex and interesting problems than we do. 





## 3) Generalizations of the three-species model and other examples

### a) Generalizations of the three-species model

The three-species model presented previously is a **simplified model** that models a perfectly linear trophic chain. However, nature is more complex and a species can be preyed upon by multiple predators and predators can predate multiple prey items.

We invite you to go see the course of Lalith Devireddy (2016) entitled "Extending the Lotka-Volterra Equations" which develops this model and its resolutions.

Many models with 3-ODE systems exist to model ecological dynamics. Here we will present two additional ones to you but we will not develop their resolution.

### b) NPZ model

The NPZ model makes it possible to **model the lake production of a pelagic ecosystem**. This model studies variations in nutrients, phytoplankton and zooplankton that are linked to each other.

Here are the equations of the system, but it will not be worked up here.

# Conclusion & motivation words!
