---
title: "Linear modelling in ecology"
bibliography: references.bib
execute: 
  freeze: auto
output: 
  html_document:
   toc: true
   toc_float: true
---

This chapter is a simple example using R

You can import R package using the code

```{r}
library(tidyverse)
```


and then describe the purpose of your chapter as well as executing R command.


For example a basic summary of a dataset is given by 

```{r}
df <- read.table("https://gist.githubusercontent.com/slopp/ce3b90b9168f2f921784de84fa445651/raw/4ecf3041f0ed4913e7c230758733948bc561f434/penguins.csv", sep = "," , header = TRUE)
```

and produce a graph

```{r}
df %>% ggplot() +
	aes(x=species, y = body_mass_g) +
	geom_boxplot()  
```


A citation @bauer2023writing

Au sein de chaque groupe, on peut créer des branches pour répartir le travail de rédaction.

Intro #Fanny

I. Modèles linéaires généraux
1) ANOVA #Clément
2) ANCOVA #Lucia


fwbdxfv 
3) Régression #Fanny


blabla bla 

4) Validation des modèles #Lucia

II. Modèles linéaires généralisés
1) Loi binômiale #Sarah
<<<<<<< HEAD

Reminders :

X a binary data with two modalities : 1 (success) et 0 (failure).
The probability of success is : $$ P(X = 1) = \pi $$
Then Y, a variable corresponding to N random draws of X, follows a binomial law :
Y ~ B(N,p)

The density function of Y is : $$ f(y;\pi)=\binom{N}{y}.\pi^{y}.(1-\pi)^{(N-y)}$$
The mean and variance of Y are : 
$$E(Y)=N.\pi\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:var(Y)=N.\pi.(1-\pi)$$

The link function is the function logit : $$logit(\mu_{y})= \alpha+ \beta_{1}.X{i1}+ \beta_{2}.X{i2}+\beta_{3}.X{i3}+...\beta_{p}.X{ip} = \eta$$
With : 
$$\mu_{y}= \frac {e^{\eta}}{1+e^{\eta}}\:\:the\:predicted\:value$$

Let's practice with an example. 


2) Loi de Poisson #Léa


test 
3) Loi négative binômiale #Léa
4) Validation des modèles #Clément

Conclusion #Sarah

Remarques :
-ici on ne parle pas de partie Exploration des données
-on ne parle pas de Gamma mais il faudra la mentionner (signaler qu'elle existe mais pas dans le programme M2)

Consignes suppl :
-didactique -> phrases courtes, claires, une phrase = une idée
-25 pages max