---
title: "Linear modelling in ecology"
bibliography: references.bib
execute: 
  freeze: auto
output: 
  html_document:
   toc: true
   toc_float: true
---

This chapter is a simple example using R

You can import R package using the code

```{r}
library(tidyverse)
```


and then describe the purpose of your chapter as well as executing R command.


For example a basic summary of a dataset is given by 

```{r}
df <- read.table("https://gist.githubusercontent.com/slopp/ce3b90b9168f2f921784de84fa445651/raw/4ecf3041f0ed4913e7c230758733948bc561f434/penguins.csv", sep = "," , header = TRUE)
```

and produce a graph

```{r}
df %>% ggplot() +
	aes(x=species, y = body_mass_g) +
	geom_boxplot()  
```


A citation @bauer2023writing

Au sein de chaque groupe, on peut créer des branches pour répartir le travail de rédaction.

Intro #Fanny
## THE STATISTICAL TEST
The purpose of statistics is to test a hypothesis. You would need to follow these steps :
###1. The null and alternative hypotheses
They define different proposed relationships between dependent and independent variable(s).
*Null hypothesis H0 = no relationship between dependent and independent variables.*
*Alternative hypothesis H1 = some expected relationship between the variables.*
###2. Calcul of the observed value Tobs
The observed value Tobs of the test statistic T is calculated from data given that the T variable has a probability distribution known under H0.
###3. The P-value
To get this value, you need to calculate the probability of observing Tobs value in this T distribution knowing the distribution of the test statistic T under H0. 
###4. Rejection or validation of the hypothesis
If the P-value is less than (or equal to) $\alpha$, then the null hypothesis is
rejected in favor of the alternative hypothesis and if the P-value is greater than $\alpha$ then the null hypothesis is not rejected.
WARNING : never say that the null hypothesis is validated. You just know that you cannot reject it with the information you have. There could be relationships that have not been detected.

## STATISTICAL MODELLING
Linear modelling is widely used in statistics to model observed data, by considering their random nature.
It explains one dependant variable, noted Y (random variable, also called **the response variable**) in function to independent variables (also called **predictors** or **explanatory variables**), also observed/measured on statistical units of the sample.
In this chapter, you will see different models corresponding to different
statistical tests.
It happens that several tests can be conducted for the same dataset and hypothesis ; in this case, you will select the most powerful test. It is the one with the lowest $\beta$-error (given at the end).

I. Modèles linéaires généraux
First let's get to know the most basic and simple linear models. We also call them *'Gaussian'* because of their allure.
They are used and validated following three eligibility conditions that you will find in part 'Model validation'.

1) ANOVA #Clément
2) ANCOVA #Lucia
3) Régression #Fanny
4) Validation des modèles #Lucia

II. Modèles linéaires généralisés

1) Loi binômiale #Sarah
2) Loi de Poisson #Léa
3) Loi négative binômiale #Léa
4) Validation des modèles #Clément

Conclusion #Sarah

Remarques :
-ici on ne parle pas de partie Exploration des données
-on ne parle pas de Gamma mais il faudra la mentionner (signaler qu'elle existe mais pas dans le programme M2)

Consignes suppl :
-didactique -> phrases courtes, claires, une phrase = une idée
-25 pages max