---
title: "Mixed models in ecology"
bibliography: references.bib
execute: 
  freeze: auto
output: 
  html_document:
   toc: true
   toc_float: true
---



## Introduction : Louis




## Model: création du model : Abdou



## Calcule matriciel : Antoine




## Structure de dependance : Odin




### Application : Ael
--> Se placer dans le contexte d'un modèle générale mixte
--> Choisir un exemple, l'expliquer, présenter les variables
### Exploration des données
--> Présenter les règles
--> Les appliquer à notre exemple
--> Sortir les graphes
--> Intepréter
### Création du model
### Validation
### Interpretation



This chapter is a simple example using R

You can import R package using the code

```{r}
library(tidyverse)
```


and then describe the purpose of your chapter as well as executing R command.


For example a basic summary of a dataset is given by 

```{r}
df <- read.table("https://gist.githubusercontent.com/slopp/ce3b90b9168f2f921784de84fa445651/raw/4ecf3041f0ed4913e7c230758733948bc561f434/penguins.csv", sep = "," , header = TRUE)
```

and produce a graph

```{r}
df %>% ggplot() +
	aes(x=species, y = body_mass_g) +
	geom_boxplot()  
```


A citation @bauer2023writing