---
title: "Mixed models in ecology"
bibliography: references.bib
execute: 
  freeze: auto
output: 
  html_document:
   toc: true
   toc_float: true
---



## Introduction : Louis

To model the relationships between a dependent variable $Y$ and one or several independent variables $X_1,\ X_2,\ ...,\ X_n$, **general linear models** (e.g. linear regressions, ANOVA, ANCOVA) are often used. In this models, several assumptions are made : homodescasticity of the variance, non-colinearity of the residuals and normality of the residuals. Homoscedasticity and normality hypothesis can be outreached using Generalized Linear Models. These models allow to use data that are drawn from different distributions such as Poisson, binomial or Gamma distributions, which are commonplace in ecology. 




## Model: création du model : Abdou



## Calcule matriciel : Antoine

$$
  M_{2\times2} = \left( {\begin{array}{cc}
    A & B \\
    C & D \\
  \end{array} } \right)
$$


## Structure de dependance : Odin




## Application : Ael
### Exploration des données
### Création du model
### Validation
### Interpretation



This chapter is a simple example using R

You can import R package using the code

```{r}
library(tidyverse)
```


and then describe the purpose of your chapter as well as executing R command.


For example a basic summary of a dataset is given by 

```{r}
df <- read.table("https://gist.githubusercontent.com/slopp/ce3b90b9168f2f921784de84fa445651/raw/4ecf3041f0ed4913e7c230758733948bc561f434/penguins.csv", sep = "," , header = TRUE)
```

and produce a graph

```{r}
df %>% ggplot() +
	aes(x=species, y = body_mass_g) +
	geom_boxplot()  
```


A citation @bauer2023writing